---
title: "AnnotatedBCGEData"
author:
  - name: "Heidi Steadman"
    affiliation:
      - &id "Brigham Young University"
  - name: "Shu-Fei Lo"
    affiliation: *id
  - name: "Stephen Piccolo"
    affiliation: *id
package: "AnnotatedBCGEData"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{AnnotatedBCGEData}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---


```{r, setup, echo=FALSE, include=FALSE}
pkg <- read.dcf("../DESCRIPTION", fields = "Package")[1]
library(pkg, character.only = TRUE)
```

# Abstract
`AnnotatedBCGEData` is an R package that uses the ExperimentHub infrastructure in [Bioconductor](https://bioconductor.org/). To create this package, we gathered more than 100 gene-expression data sets (microarray and RNA-Sequencing) from public repositories, including [Gene Expression Omnibus](http://www.ncbi.nlm.nih.gov/geo/), [ArrayExpress](https://www.ebi.ac.uk/biostudies/arrayexpress), and [The Cancer Genome Atlas](https://www.cancer.gov/ccg/research/genome-sequencing/tcga). Where feasible, we obtained raw data and reprocessed the data using modern computational pipelines. Additionally, we provide metadata in an easy-to-use format. Our goal is to provide researchers with easier access to these data with the goal of accelerating biomedical discovery.


# Installation Instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then install `AnnotatedBCGEData` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("AnnotatedBCGEData")
```

This code installs BiocManager if it is not already installed. Running the if statement is required to install the package as it is dependent on BiocManager to load. 

Download the development version from [GitHub](https://github.com/srp33/AnnotatedBCGEData) with:

```{r 'install_dev', eval = FALSE}
BiocManager::install("srp33/AnnotatedBCGEData")
```

# Data Info
The data used in this package was accessed through [Zenodo](https://zenodo.org/communities/annnotatedbcgedata/records?q=&l=list&p=1&s=10&sort=newest). This database allowed us to use data that was already neat and ready to be analyzed. However, the original data came from Gene Expression Omnibus (GEO), ArrayExpress, The Cancer Genome Atlas, METABRIC, and a repository for the International Cancer Genome Atlas that no longer exists. Learn more about the data at the following links, or via the included functions in the package:

 - GEO: https://www.ncbi.nlm.nih.gov/geo/ 
 - ArrayExpress: https://www.ebi.ac.uk/biostudies/arrayexpress 
 - TCGA: https://www.cancer.gov/ccg/research/genome-sequencing/tcga  



# Filtering Data by Ontology
``` {r load_tidyverse, include = FALSE}
library(tidyverse)
library(SummarizedExperiment)
library(zen4R)
library(BiocFileCache)
library(AnnotatedBCGEData)
```
This package contains several functions that are designed to help users filter the included data sets based on their associated ontology terms. The first function is `searchDefs`. `searchDefs` was created with the intention of helping users become more familiar with the ontology included in this package. This function takes two parameters: the term to search by and the type of term it is. The options for terms are Name, URI, Code, and Definition. If no term type is specified, the function defaults to Name. The function returns a tibble with information about the term searched, including a code unique to each term. The code can then be passed to the `searchForDatasets`function, which returns all data sets that were associated with that term. <br>
<br>
`searchForDatasets` takes two parameters: the code to search by and the type of term it is. The options are field or value. If no term is specified, it defaults to field. A field refers to the actual NCIT term code. This would be the code associated with a term such as "ER status" or "Patient age". A value refers to the particular values in the data. This could be "Positive", "Negative", or a range of values in some cases. All codes are accessible with the `searchDefs` function. `searchForDatasets` returns a tibble with all the data sets that are associated with the ontology code, as well as data set-level metadata. The names of the data sets can then be passed to another helper function to create the SummarizedExperiment object available for further analysis. The following examples show how to use the functions. 

## Example: Searching for a data set by ontology term name
In this first example of using the `searchDefs` function, we search for a particular ontology term, "Progesterone Receptor Status". 

```{r PR_example}
searchDefs("Progesterone Receptor Status")

```

The output of calling this function is a tibble containing the URI, the name of the term, synonyms, the definition, and the NCIT code associated with it. When the correct term is identified, the user can then pass the code to the `searchForDatasets` function, which identifies data sets that contain the chosen ontology term.

```{r PR_example_p2}
searchForDatasets("C16149", "field")

```

## Example: Searching for a data set by ontology term definition
In this example, we search for the ontology term by an associated definition. Because the definitions are more vague, this may identify several similar terms. It is important that the user is able to identify the exact term they are looking for.

```{r PR_example2}
searchDefs("Progesterone receptor", "Definition")

```

Again, when the chosen term is identified, it can be passed to the `searchForDatasets` function. 

```{r PR_example2_p2}
searchForDatasets("C16149", "field")

```

# Making SummarizedExperiment Objects

The following is the code required to access the objects within the package. Other packages were used in the creation of the package, but they should be automatically loaded when the package is installed. These packages are included in the NAMESPACE and DESCRIPTION files. 

```{r load_packages, message=FALSE, warning=FALSE}
library(AnnotatedBCGEData)
```

Downloading this package does not download the data sets onto the user's machine. To load the SummarizedExperiment object for a data set included in AnnotatedBCGEData, the user chooses the name of the data set and passes it as string to the function bcgeData. This function downloads the data and creates the SummarizedExperiment object that is accessible to the user. <br>
<br>
The SummarizedExperiment object for each data set includes 3 matrices. The first, which we load as expression_data, is a matrix with the samples as columns and genes as rows. The Ensembl gene ID is used as the row names. The data in the actual matrix is the expression level of the gene for the sample. <br>
<br>
The second matrix we load as sample_metadata. This is information about the samples themselves, such as demographics for the patient the sample came from. The samples are rows and the information is in the columns. <br>
<br>
The third matrix we load as feature_data. This is the information about the genes that were tested for each sample. This includes information like the chromosome the gene is located on, the Entrez gene ID, and the gene name. The genes are the rows and the information about the samples is in the columns.

## bcgeData
The `bcgeData` function returns the SummarizedExperiment object for the requested data set. The function creates a cache (using BiocFileCache) on the user's machine. All downloaded files are saved to this cache so that files do not need to be redownloaded every time the R session is restarted. By default, the most recent version of the data that is found in the Zenodo repository is downloaded. <br> 
<br>
We understand the importance of data versioning and reproducibility. For that reason, there is additional functionality included in the bcgeData function that will allow the user to specify which version of the data they want to download or use. The argument `v` to the function takes an integer that is used to find the specified version. If that version does not exist or needs to be downloaded, the user will see a message in the R console. <br>
<br>
While the actual files are stored in the AnnotatedBCGEData cache that is created, for reasons of data version checks, the SummarizedExperiment objects are not stored in the cache and need to be created again when the R session is restarted. This is a fairly quick process as the data already exists. This requires nothing more than calling `bcgeData` on the desired data set at the beginning of every session.<br>
<br>
The following is a simple example of creating an object and accessing the data.

```{r example, eval = requireNamespace('AnnotatedBCGEData')}
GSE41197_SE = bcgeData("GSE41197")
expression_data = assay(GSE41197_SE)
sample_metadata = colData(GSE41197_SE)
feature_data = rowData(GSE41197_SE)
```

## Example 1: Plotting Expression Values
In the following example, we first access the matrix of gene expression values. To make it easier to analyze, we convert it into a tibble and select the first 10 rows. We then used ggplot() to create a bar plot showing the gene expression levels for one sample across 10 genes.

```{r plot_example, message=FALSE, warning=FALSE}
GSE41197 = bcgeData("GSE41197")
gene_expression_values = assay(GSE41197)
```

```{r plot_example_part_2}
exp_tib = as_tibble(gene_expression_values, rownames='Ensembl_Gene_ID')[1:10,] %>%
    ggplot(aes(x=Ensembl_Gene_ID, y=GSM1010328)) +
    geom_col() +
    labs(x='Ensembl Gene ID',y='Sample ID GSM1010328') +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

exp_tib
```

## Example 2: Expression Box Plot
Here, we first accessed the matrices for 3 data sets included in AnnotatedBCGEData. 

```{r plot_example_2, message=FALSE, warning=FALSE}
GSE10797_SE = bcgeData("GSE10797")
GSE41197_SE = bcgeData("GSE41197")
GSE59772_SE = bcgeData("GSE59772")

GSE10197 = assay(GSE10797_SE)
GSE41197 = assay(GSE41197_SE)
GSE59772 = assay(GSE59772_SE)
```

Next, we converted each to a tibble and selected one row, representing one gene that we knew each data set had in common. We rotated the tibbles to have one column (the gene) and many rows (the sample). Then, we added the 'group' column, which was necessary for the function used in the next step.

```{r plot_example_2_part_2}
GSE10197_gene = GSE10197 %>%
    as_tibble(rownames='Ensembl_Gene_ID') %>%
    filter(Ensembl_Gene_ID == "ENSG00000171428") %>%
    pivot_longer(cols=-Ensembl_Gene_ID,names_to='Sample_ID',values_to='Expression_Level')%>%
    mutate(group = 'GSE10197')
GSE41197_gene = GSE41197 %>%
    as_tibble(rownames='Ensembl_Gene_ID') %>%
    filter(Ensembl_Gene_ID == "ENSG00000171428") %>%
    pivot_longer(cols=-Ensembl_Gene_ID,names_to='Sample_ID',values_to='Expression_Level')%>%
    mutate(group = 'GSE41197')
GSE59772_gene = GSE59772 %>%
    as_tibble(rownames='Ensembl_Gene_ID') %>%
    filter(Ensembl_Gene_ID == "ENSG00000171428") %>%
    pivot_longer(cols=-Ensembl_Gene_ID,names_to='Sample_ID',values_to='Expression_Level')%>%
    mutate(group = 'GSE59772')
```

We then joined the 3 tibbles into one tibble with the bind_rows() function. We now have one tibble with one column (the gene) and many rows (the samples). <br>
<br>
Finally, we used ggplot() to visualize the expression levels for the gene across the three data sets.

```{r plot_example_2_part_3}
combined_samples = bind_rows(GSE10197_gene, GSE41197_gene, GSE59772_gene)

ggplot(combined_samples, aes(x = group, y = Expression_Level, fill = group)) +
    geom_boxplot() +
    labs(x = "Dataset", y = "Expression Level") +
    theme_bw()
```

## Example 3: Using Metadata
In this example, we combine using both the metadata and the expression data to create a heat map of gene expression levels.<br>
First, we load the data sets, turn them into tibbles, and extract both the metadata and expression data. We then filter the expression to only include the samples that are identified as stroma cells. Finally, we create two vectors with all the gene names in each tibble to set them up for comparison. 

``` {r heat_map, message=FALSE, warning=FALSE}
GSE59772_metadata = colData(GSE59772_SE) %>%
    as_tibble(rownames = 'Sample_ID')

GSE59772_stroma_samples = filter(GSE59772_metadata, tissue == 'Stroma') %>%
    pull(Sample_ID)

GSE59772_gene_names = assay(GSE59772_SE) %>%
    as_tibble(rownames="Ensembl_Gene_ID") %>%
    pull(Ensembl_Gene_ID)


GSE10797 = bcgeData("GSE10797")
GSE10797_metadata = colData(GSE10797) %>%
    as_tibble(rownames = 'Sample_ID')

GSE10797_gene_names = assay(GSE10797) %>%
    as_tibble(rownames="Ensembl_Gene_ID") %>%
    pull(Ensembl_Gene_ID)

GSE10797_stroma_samples = filter(GSE10797_metadata, tissue_source == 'stromal cells from breast cancer patient')%>%
    pull(Sample_ID)
```

Next, we create a vector that has only the names of the genes the data sets have in common. We use this vector to filter the expression tibbles again to only include rows with genes in the vector. 

``` {r heat_map_2}
common_genes = c()
for (gene in GSE59772_gene_names) {
    if (gene %in% GSE10797_gene_names) {
        common_genes = c(common_genes, gene)
    }
}

GSE59772_stroma_expressions = assay(GSE59772_SE) %>%
    as_tibble(rownames= 'Ensembl_Gene_ID') %>%
    select(Ensembl_Gene_ID, all_of(GSE59772_stroma_samples)) %>%
    filter(Ensembl_Gene_ID %in% common_genes)
    
GSE10797_stroma_expressions = assay(GSE10797) %>%
    as_tibble(rownames = 'Ensembl_Gene_ID') %>%
    select(Ensembl_Gene_ID, all_of(GSE10797_stroma_samples)) %>% 
    filter(Ensembl_Gene_ID %in% common_genes)

```

Finally, we combine the data into one tibble called combined_data. We make a new row called variance that calculates the variance across the 5 samples. We select the top 10 rows, those with greatest variance. We then use ggplot to create a heat map of the genes with the top variance, displayed below. 

``` {r heat_map_3}
combined_data = full_join(GSE10797_stroma_expressions, GSE59772_stroma_expressions)
top_variance_genes = combined_data %>%
    rowwise() %>%
    mutate(variance = var(c_across(starts_with('GSM')))) %>%
    ungroup() %>%
    arrange(desc(variance)) %>%
    slice_head(n=10)

top_variance_map = top_variance_genes %>%
    select(-variance) %>%
    pivot_longer(cols = starts_with('GSM'), names_to = 'Sample_ID', values_to = 'Expression') %>%
    ggplot(aes(x = Sample_ID, y = Ensembl_Gene_ID, fill = Expression)) +
    geom_tile()+
    labs(x = 'Sample', y = 'Gene', fill = 'Expression', title = 'Top 10 Genes in Stroma Samples with Highest Variance') + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1), plot.title = element_text(hjust = 0.5, face = "bold"))
top_variance_map


```

## Example 4: Using Feature Data
In this final example, we use aspects of the feature data along with the expression data. <br> <br>

First, we get the feature data and convert it to a tibble. We then filter this data for each data set to only include genes on chromosome 16. Similar to the last example, we also make a vector with the gene names the data sets have in common. 

``` {r chr16_boxplot, warning=FALSE}
GSE59772_featuredata = rowData(GSE59772_SE) %>%
    as_tibble(rownames = 'Ensembl_Gene_ID')

GSE59772_chr16 = GSE59772_featuredata %>%
    filter(Chromosome == '16') %>%
    pull(Ensembl_Gene_ID)

GSE59772_gene_names = assay(GSE59772_SE) %>%
    as_tibble(rownames="Ensembl_Gene_ID") %>%
    pull(Ensembl_Gene_ID)


GSE10797_featuredata = rowData(GSE10797) %>%
    as_tibble(rownames = 'Ensembl_Gene_ID')

GSE10797_chr16 = GSE10797_featuredata %>%
    filter(Chromosome == '16') %>%
    pull(Ensembl_Gene_ID)

GSE10797_gene_names = assay(GSE10797) %>%
    as_tibble(rownames="Ensembl_Gene_ID") %>%
    pull(Ensembl_Gene_ID)


common_genes = c()
for (gene in GSE59772_gene_names) {
    if (gene %in% GSE10797_gene_names) {
        common_genes = c(common_genes, gene)
    }
}
```

Next, we filter the expression data to include only genes on chromosome 16, then filter again to only include genes the data sets have in common. We pivot and mutate to make the data ready for the boxplot. 
``` {r chr16_boxplot_2}
GSE10797_chr16_common = assay(GSE10797) %>%
    as_tibble(rownames = 'Ensembl_Gene_ID') %>%
    filter(Ensembl_Gene_ID %in% GSE10797_chr16) %>%
    filter(Ensembl_Gene_ID %in% common_genes) %>%
    pivot_longer(cols=-Ensembl_Gene_ID,names_to='Sample_ID',values_to='Expression_Level')%>%
    mutate(Dataset = 'GSE10797')

GSE59772_chr16_common = assay(GSE59772_SE) %>%
    as_tibble(rownames = 'Ensembl_Gene_ID') %>%
    filter(Ensembl_Gene_ID %in% GSE59772_chr16) %>%
    filter(Ensembl_Gene_ID %in% common_genes) %>%
    pivot_longer(cols=-Ensembl_Gene_ID,names_to='Sample_ID',values_to='Expression_Level')%>%
    mutate(Dataset = 'GSE59772')
```

Finally, we combine the data sets using bind_rows() and read it into a ggplot. 
``` {r chr16_boxplot_3}
combined_data = bind_rows(GSE10797_chr16_common, GSE59772_chr16_common) %>%
    ggplot(aes(x = Dataset, y = Expression_Level, fill = Dataset)) +
    geom_boxplot() +
    theme_bw() +
    labs(x = 'Dataset', y='Gene Expression Level', title = 'Gene Expression Levels in Chromosome 16') 
combined_data
```

# Session Info 

<details> 

```{r Session Info}
utils::sessionInfo()
```

</details>  

<br>
