---
title: "AnnotatedBCGEData"
author:
  - name: "Heidi Steadman"
    affiliation:
      - &id "Brigham Young University"
  - name: "Shu-Fei Lo"
    affiliation: *id
  - name: "Stephen Piccolo"
    affiliation: *id
package: "AnnotatedBCGEData"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{AnnotatedBCGEData}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, setup, echo=FALSE, include=FALSE}
pkg <- read.dcf("../DESCRIPTION", fields = "Package")[1]
library(pkg, character.only = TRUE)
```

# Abstract

`AnnotatedBCGEData` is an R/[Bioconductor](https://bioconductor.org) package that provides access to more than 100 gene-expression datasets (generated using microarray and RNA-Sequencing technologies). We collected these datasets from public repositories, including [Gene Expression Omnibus](http://www.ncbi.nlm.nih.gov/geo/), [ArrayExpress](https://www.ebi.ac.uk/biostudies/arrayexpress), and [The Cancer Genome Atlas](https://www.cancer.gov/ccg/research/genome-sequencing/tcga). Where feasible, we obtained and reprocessed the raw data using computational pipelines and annotations that are standardized and more updated than those used to process the data originally. Additionally, we annotated the sample metadata using ontology terms from the [National Cancer Institute Thesaurus](https://evsexplore.semantics.cancer.gov/evsexplore/welcome). This was necessary because different data submitters used different terminology to describe the samples. Mapping the metadata to ontology terms makes it possible for users to search our data using these standardized terms.

# Installation instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/).

Then install `AnnotatedBCGEData` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r install, eval = FALSE}
## TODO: need to verify that this code is accurate for downloading from Bioconductor when it is available
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("AnnotatedBCGEData")
```

Download the development version from [GitHub](https://github.com/srp33/AnnotatedBCGEData) with:

```{r install_dev, eval = FALSE}
BiocManager::install("srp33/AnnotatedBCGEData")
```

Finally, to load the package, the following code is required:

```{r load_example, eval = FALSE}
library(AnnotatedBCGEData)
```


# Data Persistence
Attempting to store all of the data within this package would not be feasible due to the data's size. Thus, we are storing the data remotely and accessing it when requested. The data can optionally be stored in a persistent cache on the user's machine if desired. Once a user has identified dataset(s) to work with, our package retrieves the data from a public [Zenodo](https://zenodo.org/communities/annnotatedbcgedata/records) repository. If they prefer, users can access the files directly on Zenodo in TSV format. They can also find details about the data by exploring our Zenodo repository. <br><br>

The package will allow the persistent storage of data across R sessions if the user specifies a filepath to save the data to when calling `getBCGEData`. If no filepath is specified, the data will be saved to a temporary directory that will refresh when the R session is restarted. <br><br>

It should be noted that at the time of package development, Zenodo API limitations reduce the number of requests per minute to 30. Because of this, it has been found necessary to allow breaks between successive calls of `getBCGEData`. The API limitations should not cause problems for the user, but users should be aware of an increased amount of time when downloading large amounts of data.

# Filtering datasets by ontology terms

```{r load_packages, include = FALSE}
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyr)
library(zen4R)
library(AnnotatedBCGEData)
library(SummarizedExperiment)
```

This package contains functions designed to help users filter the included data sets based on their associated ontology terms. The first function is `searchOntologyTerms`. `searchOntologyTerms` was created with the intention of helping users become more familiar with the ontology included in this package. This function takes two parameters: the term to search by and the type of term it is. The term to search by is specified as a string; either the term itself, a code, a URI, or a definition/description. The options for terms are Name, URI, Code, and Definition, also passed as a string. If no term type is specified, the function defaults to Name. The function returns a tibble with information about the term searched, including a code unique to each term. The code can then be passed to the `searchForDatasets`function, which returns all data sets that were associated with that term. <br>
<br>
`searchForDatasets` takes two parameters: the code to search by and the type of term it is. The options are field or value. If no term is specified, it defaults to field. A field refers to the actual NCIT term code. This would be the code associated with a term such as "ER status" or "Patient age". A value refers to the particular values in the data. This could be "Positive", "Negative", or a range of values in some cases. All codes are accessible with the `searchOntologyTerms` function. `searchForDatasets` returns a tibble with all the data sets that are associated with the ontology code, as well as data set-level metadata. The names of the data sets can then be passed to another helper function to create the SummarizedExperiment object available for further analysis. The following examples show how to use the functions. 

## Example: Searching for a data set by ontology term name
In this first example of using the `searchOntologyTerms` function, we search for a particular ontology term, "Progesterone Receptor Status". 

```{r PR_example}
defs <- searchOntologyTerms("Progesterone Receptor Status")
```

The output of calling this function is a tibble containing the URI, the name of the term, synonyms, the definition, and the NCIT code associated with it. When the correct term is identified, the user can then pass the code to the `searchForDatasets` function, which identifies data sets that contain the chosen ontology term.

```{r PR_example_p2}
datasets <- searchForDatasets("C16149", "field")
```

## Example: Searching for a data set by ontology term definition
In this example, we search for the ontology term by an associated definition. Because the definitions are more vague, this may identify several similar terms. It is important that the user is able to identify the exact term they are looking for.

```{r PR_example2}
searchOntologyTerms("Progesterone receptor", "Definition")

```

Again, when the chosen term is identified, it can be passed to the `searchForDatasets` function. When desired datasets are found, they can then be passed to `getBCGEData` to download the data and access the SummarizedExperiment object.

```{r PR_example2_p2}
searchForDatasets("C16149", "field")
```

# Making SummarizedExperiment Objects

Downloading this package does not download the data sets onto the user's machine. To load the SummarizedExperiment object for a data set included in AnnotatedBCGEData, the user chooses the name of the data set and passes it as string to the function getBCGEData. This function downloads the data and creates the SummarizedExperiment object that is accessible to the user. <br>
<br>
The SummarizedExperiment object for each data set includes 3 matrices. The first, which we load as expression_data, is a matrix with the samples as columns and genes as rows. The Ensembl gene ID is used as the row names. The data in the actual matrix is the expression level of the gene for the sample. <br>
<br>
The second matrix we load as sample_metadata. This is information about the samples themselves, such as demographics for the patient the sample came from. The samples are rows and the information is in the columns. <br>
<br>
The third matrix we load as feature_data. This is the information about the genes that were tested for each sample. This includes information like the chromosome the gene is located on, the Entrez gene ID, and the gene name. The genes are the rows and the information about the samples is in the columns.

## getBCGEData
The `getBCGEData` function returns the SummarizedExperiment object for the requested data set. The function creates a cache (using BiocFileCache) on the user's machine. All downloaded files are saved to this cache so that files do not need to be redownloaded every time the R session is restarted. By default, the most recent version of the data that is found in the Zenodo repository is downloaded. When called for the first time, the user will be prompted to select or enter a location for the cache to be saved.<br> 
<br>
We understand the importance of data versioning and reproducibility. For that reason, there is additional functionality included in the getBCGEData function that will allow the user to specify which version of the data they want to download or use. The argument `v` to the function takes an integer that is used to find the specified version. If that version does not exist or needs to be downloaded, the user will see a message in the R console. If `v` is not specified, the most recent version is downloaded and used. <br>
<br>
While the actual files are stored in the AnnotatedBCGEData cache that is created, for reasons of data version checks, the SummarizedExperiment objects are not stored in the cache and need to be created again when the R session is restarted. This is a fairly quick process as the data already exists. This requires nothing more than calling `getBCGEData` on the desired data set at the beginning of every session.<br>
<br>
The following is a simple example of creating an object and accessing the data.

```{r example, eval = requireNamespace('AnnotatedBCGEData')}
GSE41197_SE = getBCGEData("GSE41197")
GSE41197_expression_data = assay(GSE41197_SE)
sample_metadata = colData(GSE41197_SE)
feature_data = rowData(GSE41197_SE)
```

Similarly, if the user wanted to access an older version of the same dataset, the function could be used as follows;
``` {r version_example}
GSE41197_v2 = getBCGEData("GSE41197", v=2)
GSE41197_v2_expressions = assay(GSE41197_v2)
v2_metadata = colData(GSE41197_v2)
v2_feature_data = rowData(GSE41197_v2)

```
## Example 1: Plotting Expression Values
In the following example, we demonstrate data exploration that can be done easily with the package. We first access the matrix of gene expression values. To make it easier to analyze, we convert it into a tibble and select the first 10 rows. We then used ggplot() to create a bar plot showing the gene expression levels for one sample across 10 genes.

```{r plot_example_part_2}
exp_tib = as_tibble(GSE41197_expression_data, 
                    rownames='Ensembl_Gene_ID')[1:10,] %>%
    ggplot(aes(x=Ensembl_Gene_ID, y=GSM1010328)) +
    geom_col() +
    labs(x='Ensembl Gene ID',y='Sample ID GSM1010328') +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

exp_tib
```


## Example 2: Using Metadata
In this example, we combine using both the metadata and the expression data to create a heat map of gene expression levels.<br>
First, we load the data sets, turn them into tibbles, and extract both the metadata and expression data. We then filter the expression to only include the samples that are identified as stroma cells. Finally, we create two vectors with all the gene names in each tibble to set them up for comparison. 

``` {r heat_map, message=FALSE, warning=FALSE}
GSE10797_SE <- getBCGEData("GSE10797")
GSE59772_SE <- getBCGEData("GSE59772")


GSE59772_exp <- assay(GSE59772_SE)
GSE10797_exp <- assay(GSE10797_SE)

GSE59772_metadata = colData(GSE59772_SE) %>%
    as_tibble(rownames = 'Sample_ID')

GSE59772_stroma_samples = filter(GSE59772_metadata, tissue == 'Stroma') %>%
    pull(Sample_ID)

GSE59772_gene_names = GSE59772_exp %>%
    as_tibble(rownames="Ensembl_Gene_ID") %>%
    pull(Ensembl_Gene_ID)


GSE10797_metadata = colData(GSE10797_SE) %>%
    as_tibble(rownames = 'Sample_ID')

GSE10797_gene_names = GSE10797_exp %>%
    as_tibble(rownames="Ensembl_Gene_ID") %>%
    pull(Ensembl_Gene_ID)

GSE10797_stroma_samples = filter(GSE10797_metadata, 
                tissue_source == 'stromal cells from breast cancer patient')%>%
    pull(Sample_ID)
```

Next, we create a vector that has only the names of the genes the data sets have in common. We use this vector to filter the expression tibbles again to only include rows with genes in the vector. 

``` {r heat_map_2}
common_genes = c()
for (gene in GSE59772_gene_names) {
    if (gene %in% GSE10797_gene_names) {
        common_genes = c(common_genes, gene)
    }
}

GSE59772_stroma_expressions = GSE59772_exp %>%
    as_tibble(rownames= 'Ensembl_Gene_ID') %>%
    select(Ensembl_Gene_ID, all_of(GSE59772_stroma_samples)) %>%
    filter(Ensembl_Gene_ID %in% common_genes)
    
GSE10797_stroma_expressions = GSE10797_exp %>%
    as_tibble(rownames = 'Ensembl_Gene_ID') %>%
    select(Ensembl_Gene_ID, all_of(GSE10797_stroma_samples)) %>% 
    filter(Ensembl_Gene_ID %in% common_genes)

```

Finally, we combine the data into one tibble called combined_data. We make a new row called variance that calculates the variance across the 5 samples. We select the top 10 rows, those with greatest variance. We then use ggplot to create a heat map of the genes with the top variance, displayed below. 

``` {r heat_map_3}
combined_data = full_join(GSE10797_stroma_expressions, 
                          GSE59772_stroma_expressions)
top_variance_genes = combined_data %>%
    rowwise() %>%
    mutate(variance = var(c_across(starts_with('GSM')))) %>%
    ungroup() %>%
    arrange(desc(variance)) %>%
    slice_head(n=10)

top_variance_map = top_variance_genes %>%
    select(-variance) %>%
    pivot_longer(cols = starts_with('GSM'), 
                 names_to = 'Sample_ID', 
                 values_to = 'Expression') %>%
    ggplot(aes(x = Sample_ID, y = Ensembl_Gene_ID, fill = Expression)) +
    geom_tile()+
    labs(x = 'Sample', y = 'Gene', fill = 'Expression', 
         title = 'Top 10 Genes in Stroma Samples with Highest Variance') + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1), 
          plot.title = element_text(hjust = 0.5, face = "bold"))
top_variance_map

```

## Example 4: Using Feature Data
In this final example, we use aspects of the feature data along with the expression data. <br> <br>

First, we get the feature data and convert it to a tibble. We then filter this data for each data set to only include genes on chromosome 16. Similar to the last example, we also make a vector with the gene names the data sets have in common. 

``` {r chr16_boxplot, warning=FALSE}
GSE59772_featuredata = rowData(GSE59772_SE) %>%
    as_tibble(rownames = 'Ensembl_Gene_ID')

GSE59772_chr16 = GSE59772_featuredata %>%
    filter(Chromosome == '16') %>%
    pull(Ensembl_Gene_ID)

GSE59772_gene_names = GSE59772_exp %>%
    as_tibble(rownames="Ensembl_Gene_ID") %>%
    pull(Ensembl_Gene_ID)


GSE10797_featuredata = rowData(GSE10797_SE) %>%
    as_tibble(rownames = 'Ensembl_Gene_ID')

GSE10797_chr16 = GSE10797_featuredata %>%
    filter(Chromosome == '16') %>%
    pull(Ensembl_Gene_ID)

GSE10797_gene_names = GSE10797_exp %>%
    as_tibble(rownames="Ensembl_Gene_ID") %>%
    pull(Ensembl_Gene_ID)


common_genes = c()
for (gene in GSE59772_gene_names) {
    if (gene %in% GSE10797_gene_names) {
        common_genes = c(common_genes, gene)
    }
}
```

Next, we filter the expression data to include only genes on chromosome 16, then filter again to only include genes the data sets have in common. We pivot and mutate to make the data ready for the boxplot. 
``` {r chr16_boxplot_2}
GSE10797_chr16_common = GSE10797_exp %>%
    as_tibble(rownames = 'Ensembl_Gene_ID') %>%
    filter(Ensembl_Gene_ID %in% GSE10797_chr16) %>%
    filter(Ensembl_Gene_ID %in% common_genes) %>%
    pivot_longer(cols=-Ensembl_Gene_ID,
                 names_to='Sample_ID',
                 values_to='Expression_Level')%>%
    mutate(Dataset = 'GSE10797')

GSE59772_chr16_common = GSE59772_exp %>%
    as_tibble(rownames = 'Ensembl_Gene_ID') %>%
    filter(Ensembl_Gene_ID %in% GSE59772_chr16) %>%
    filter(Ensembl_Gene_ID %in% common_genes) %>%
    pivot_longer(cols=-Ensembl_Gene_ID,
                 names_to='Sample_ID',
                 values_to='Expression_Level')%>%
    mutate(Dataset = 'GSE59772')
```

Finally, we combine the data sets using bind_rows() and read it into a ggplot. 
``` {r chr16_boxplot_3}
combined_data = bind_rows(GSE10797_chr16_common, GSE59772_chr16_common) %>%
    ggplot(aes(x = Dataset, y = Expression_Level, fill = Dataset)) +
    geom_boxplot() +
    theme_bw() +
    labs(x = 'Dataset', y='Gene Expression Level', 
         title = 'Gene Expression Levels in Chromosome 16') 
combined_data
```


# Session Info 

<details> 

```{r Session Info}
utils::sessionInfo()
```

</details>  

<br>
