---
title: "AnnotatedBCGEData"
author:
  - name: "Heidi Steadman"
    affiliation:
      - &id "Brigham Young University"
  - name: "Shu-Fei Lo"
    affiliation: *id
  - name: "Stephen Piccolo"
    affiliation: *id
package: "AnnotatedBCGEData"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{AnnotatedBCGEData}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, setup, echo=FALSE, include=FALSE}
pkg <- read.dcf("../DESCRIPTION", fields = "Package")[1]
library(pkg, character.only = TRUE)
```

# Abstract

`AnnotatedBCGEData` is an [R](https://r-project.org)/[Bioconductor](https://bioconductor.org) package that provides access to more than 100 gene-expression datasets generated using microarray and RNA-Sequencing technologies. We collected these datasets from public repositories, including [Gene Expression Omnibus](http://www.ncbi.nlm.nih.gov/geo/), [ArrayExpress](https://www.ebi.ac.uk/biostudies/arrayexpress), and [The Cancer Genome Atlas](https://www.cancer.gov/ccg/research/genome-sequencing/tcga). Where feasible, we obtained and reprocessed the raw data using computational pipelines and annotations that are standardized and more updated than those used to process the data originally. Additionally, we annotated the sample metadata using ontology terms from the [National Cancer Institute Thesaurus](https://evsexplore.semantics.cancer.gov/evsexplore/welcome). This was necessary because different data submitters used different terminology to describe breast-cancer patients and samples. Mapping the metadata to ontology terms makes it possible for users to search the data using these standardized terms.

# Installation instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/).

Then install `AnnotatedBCGEData` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("AnnotatedBCGEData")
```

Or install the development version:

```{r install_dev, eval = FALSE}
BiocManager::install("srp33/AnnotatedBCGEData")
```

Finally, load the package:

```{r load_example, eval = FALSE}
library(AnnotatedBCGEData)
```

# Data retrieval

Attempting to store all of the data within this package would be infeasible due to the data's size. Thus, we store the data remotely. Once a user has identified dataset(s) to analyze, our package retrieves the data from a public [Zenodo](https://zenodo.org/communities/annnotatedbcgedata/records) repository. (If they prefer, researchers can find details about the data and access the files directly by exploring our Zenodo repository.)

## Searching for ontology terms

```{r load_packages, include = FALSE}
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyr)
library(zen4R)
library(AnnotatedBCGEData)
library(SummarizedExperiment)
```

This package contains functions designed to help users filter the included datasets based on their associated ontology terms. The first function that researchers typically use is `searchOntologyTerms`. Its purpose is to help users identify ontology terms mapped to the data. It accepts two parameters: a string to search by and what should be searched. The second parameter allows the user to indicate whether the string is expected to match the "Name" (default), "Definition", "URI", or "Code" of the ontology term(s). The function returns a tibble with information about the matching terms. The user reviews the matching term(s) and retrieves the corresponding ontology code(s), which can then be passed to the `searchForDatasetsByField` function or `searchForDatasetsByValue` function, which return information about datasets associated with the term(s).

In the following example, we are using the `searchOntologyTerms` function to search for ontology terms that match the string, "Progesterone Receptor Status". Because we did not specify a value for the `term_type` argument, the default of "Name" is used.

```{r PR_example}
searchOntologyTerms("Progesterone Receptor Status")
```

In this similar example, we are searching for ontology terms that match the shorter (and lowercase) string, "receptor status".

```{r RS_example}
searchOntologyTerms("receptor status")
```

The function returns a tibble containing details about ontology terms related to both progesterone receptor status and estrogen receptor status.

The National Cancer Institute Thesaurus provides synonyms for most terms. When a user searches by "Name" with the `searchOntologyTerms` function, it automatically checks against synonyms as well. This is helpful when the search text does not match the "preferred name" of an ontology term but instead matches a synonym. For example, we could use the following code to search for ontology terms related to progesterone receptor status using a synonym that is commonly used in medical contexts.

```{r P_example}
searchOntologyTerms("PgR Status")
```

Another alternative is to search ontology term *definitions*. Searching by definitions is helpful when you don’t know the exact term but understand the concept you’re looking for. In this example, we are searching for definitions matching the string, "immunohistochemical". 

```{r P_example}
searchOntologyTerms("immunohistochemical", term_type = "Definition")
```

In this case, the function returns a tibble containing details about ontology terms for concepts that use immunohistochemical methods for profiling. This includes terms related to progesterone receptor status as well as other cell-surface markers.

## Identifying datasets associated with specific ontology terms

After we have identified ontology terms for the research topic we wish to study, we can identify datasets associated with those terms. The `searchForDatasetsByField` function accepts the code for an ontology term to search by. This code is the unique identifier associated with an ontology terms (typically found using the `searchOntologyTerms` function). The `searchForDatasetsByField` function checks the ontology terms associated with metadata columns (fields) returns a tibble with the identifier(s) of the associated dataset(s), the terms that were searched, and the associated codes. The  dataset identifiers can then be passed to the `getDataset` function to retrieve the data.

Similarly, the `searchForDatasetsByValue` function accepts the code for an ontology term that maps to values within a given column (field). The function searches across the datasets and returns a tibble with information about the matching datasets. The dataset identifiers can then be passed to the `getDataset` function to retrieve the data.

In the previous examples, we searched for ontology terms related to progesterone receptor status. One of the matching terms had a name of "Progesterone Receptor Status" and a code of "C16149". We also identified "C16150" as the code associated with the "Estrogen Receptor Status" ontology term. In the following example, we are using the `searchForDatasetsByField` function to search for datasets with a column (field) that has been mapped to either of these ontology terms. This function returns a tibble indicating the dataset identifier, 

```{r PR_example_p2}
searchForDatasetsByField(c("C16149", "C16150"))
```

Alternatively (or instead), we can use the `searchForDatasetsByValue` function to search for the datasets with values in any column that have been associated with a specified ontology term. For example, if we wanted to identify datasets with samples that had been indicated as positive for the progesterone receptor, we could use the code "C15496" for the Progesterone Receptor Positive term.

```{r PR_example_p3}
searchForDatasetsByValue("C15496")
```

## Obtaining dataset-level metadata

In addition to the ontology, users can decide which datasets to analyze based on metadata about the datasets. This is possible using the `getDatasetMetadata` function, which accepts a vector of dataset identifiers. Below is an example.

```{r get_datameta}
getDatasetMetadata(c("GSE41197", "GSE59772", "GSE10797"))
```

# Obtaining gene-expression data (as SummarizedExperiment objects)

Installing our package does not download the datasets themselves. The data files are stored in a Zenodo repository and retrieved on demand. To obtain data, use the getDataset function, which stores the data and metadata in a SummarizedExperiment object.
<br>
<br>
The SummarizedExperiment object for each dataset includes 3 items. One is the gene-expression data, a matrix with samples as columns and genes as rows. Ensembl gene identifiers are used as row names. Each data value is the expression level for its respective sample and gene. Second is the sample metadata, such as patient demographics and sample details. Each row represents a sample, and each column represents a different metadata variable. Third is feature data—information about the genes that were profiled for each sample. Each row represents a gene, providing information like the Entrez gene ID, gene symbol, and chromosome identifier.

```{r example, eval = requireNamespace('AnnotatedBCGEData')}
GSE41197_SE = getDataset("GSE41197")

expression_data = assay(GSE41197_SE)
sample_metadata = colData(GSE41197_SE)
feature_data = rowData(GSE41197_SE)
```

The `getDataset` function creates a cache (using BiocFileCache) on the user's machine. All downloaded files are saved to this cache so that files do not need to be re-downloaded every time an R session is restarted. If a new R session is started and the cached files are still present, the SummarizedExperiment objects are recreated using the cached files. The `cacheDirPath` parameter allows the user to customize this location.

In some cases, multiple versions of a given dataset are available on Zenodo. The `getDataset` function allows the user to specify which version of the data to use. By default, the most recent version of the data is downloaded. The `v` argument accepts an integer specifying the version. If that version does not exist or needs to be downloaded, a message will be displayed.
<br>
<br>

The following example illustrates retrieving a specific version of a dataset.

``` {r version_example}
GSE41197_v2 = getDataset("GSE41197", v=2)

v2_expression_data = assay(GSE41197_v2)
v2_metadata = colData(GSE41197_v2)
v2_feature_data = rowData(GSE41197_v2)
```

The `getDataset` function accepts a vector that can include multiple dataset identifiers. However, at the time of package development, the Zenodo API limits the number of requests per minute to 30. Because of this, we found it necessary to allow breaks between successive calls to Zenodo (performed by the `getDataset` function). These API limits should not cause problems for the user, but users should be aware that this may cause delays when downloading large amounts of data.

## Example 1: Plotting Expression Values

In the following example, we demonstrate examples of data exploration that can be done easily with the package. We first access the matrix of gene-expression values. To make it easier to analyze, we convert it into a tibble and select the first 10 rows. We then used ggplot() to create a bar plot showing the gene expression levels for one sample across 10 genes.

```{r plot_example_part_2}
exp_tib = as_tibble(expression_data, 
                    rownames='Ensembl_Gene_ID')[1:10,] %>%
    ggplot(aes(x=Ensembl_Gene_ID, y=GSM1010328)) +
    geom_col() +
    labs(x='Ensembl Gene ID',y='Sample ID GSM1010328') +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

exp_tib
```

## Example 2: Using Metadata

In this example, we combine using both the metadata and the expression data to create a heat map of gene expression levels.<br>
First, we load the data sets, turn them into tibbles, and extract both the metadata and expression data. We then filter the expression to only include the samples that are identified as stroma cells. Finally, we create two vectors with all the gene names in each tibble to set them up for comparison. 

``` {r heat_map, message=FALSE, warning=FALSE}
GSE10797_SE <- getDataset("GSE10797")
GSE59772_SE <- getDataset("GSE59772")


GSE59772_exp <- assay(GSE59772_SE)
GSE10797_exp <- assay(GSE10797_SE)

GSE59772_metadata = colData(GSE59772_SE) %>%
    as_tibble(rownames = 'Sample_ID')

GSE59772_stroma_samples = filter(GSE59772_metadata, tissue == 'Stroma') %>%
    pull(Sample_ID)

GSE59772_gene_names = GSE59772_exp %>%
    as_tibble(rownames="Ensembl_Gene_ID") %>%
    pull(Ensembl_Gene_ID)


GSE10797_metadata = colData(GSE10797_SE) %>%
    as_tibble(rownames = 'Sample_ID')

GSE10797_gene_names = GSE10797_exp %>%
    as_tibble(rownames="Ensembl_Gene_ID") %>%
    pull(Ensembl_Gene_ID)

GSE10797_stroma_samples = filter(GSE10797_metadata, 
                tissue_source == 'stromal cells from breast cancer patient')%>%
    pull(Sample_ID)
```

Next, we create a vector that has only the names of the genes the data sets have in common. We use this vector to filter the expression tibbles again to only include rows with genes in the vector. 

``` {r heat_map_2}
common_genes = c()
for (gene in GSE59772_gene_names) {
    if (gene %in% GSE10797_gene_names) {
        common_genes = c(common_genes, gene)
    }
}

GSE59772_stroma_expressions = GSE59772_exp %>%
    as_tibble(rownames= 'Ensembl_Gene_ID') %>%
    select(Ensembl_Gene_ID, all_of(GSE59772_stroma_samples)) %>%
    filter(Ensembl_Gene_ID %in% common_genes)
    
GSE10797_stroma_expressions = GSE10797_exp %>%
    as_tibble(rownames = 'Ensembl_Gene_ID') %>%
    select(Ensembl_Gene_ID, all_of(GSE10797_stroma_samples)) %>% 
    filter(Ensembl_Gene_ID %in% common_genes)
```

Finally, we combine the data into one tibble called combined_data. We make a new row called variance that calculates the variance across the 5 samples. We select the top 10 rows, those with greatest variance. We then use ggplot to create a heat map of the genes with the top variance, displayed below. 

``` {r heat_map_3}
combined_data = full_join(GSE10797_stroma_expressions, 
                          GSE59772_stroma_expressions)
top_variance_genes = combined_data %>%
    rowwise() %>%
    mutate(variance = var(c_across(starts_with('GSM')))) %>%
    ungroup() %>%
    arrange(desc(variance)) %>%
    slice_head(n=10)

top_variance_map = top_variance_genes %>%
    select(-variance) %>%
    pivot_longer(cols = starts_with('GSM'), 
                 names_to = 'Sample_ID', 
                 values_to = 'Expression') %>%
    ggplot(aes(x = Sample_ID, y = Ensembl_Gene_ID, fill = Expression)) +
    geom_tile()+
    labs(x = 'Sample', y = 'Gene', fill = 'Expression', 
         title = 'Top 10 Genes in Stroma Samples with Highest Variance') + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1), 
          plot.title = element_text(hjust = 0.5, face = "bold"))
top_variance_map
```

## Example 4: Using Feature Data

In this final example, we use aspects of the feature data along with the expression data. <br> <br>

First, we get the feature data and convert it to a tibble. We then filter this data for each data set to only include genes on chromosome 16. Similar to the last example, we also make a vector with the gene names the data sets have in common. 

``` {r chr16_boxplot, warning=FALSE}
GSE59772_featuredata = rowData(GSE59772_SE) %>%
    as_tibble(rownames = 'Ensembl_Gene_ID')

GSE59772_chr16 = GSE59772_featuredata %>%
    filter(Chromosome == '16') %>%
    pull(Ensembl_Gene_ID)

GSE59772_gene_names = GSE59772_exp %>%
    as_tibble(rownames="Ensembl_Gene_ID") %>%
    pull(Ensembl_Gene_ID)


GSE10797_featuredata = rowData(GSE10797_SE) %>%
    as_tibble(rownames = 'Ensembl_Gene_ID')

GSE10797_chr16 = GSE10797_featuredata %>%
    filter(Chromosome == '16') %>%
    pull(Ensembl_Gene_ID)

GSE10797_gene_names = GSE10797_exp %>%
    as_tibble(rownames="Ensembl_Gene_ID") %>%
    pull(Ensembl_Gene_ID)

common_genes = c()
for (gene in GSE59772_gene_names) {
    if (gene %in% GSE10797_gene_names) {
        common_genes = c(common_genes, gene)
    }
}
```

Next, we filter the expression data to include only genes on chromosome 16, then filter again to only include genes the data sets have in common. We pivot and mutate to make the data ready for the boxplot.

``` {r chr16_boxplot_2}
GSE10797_chr16_common = GSE10797_exp %>%
    as_tibble(rownames = 'Ensembl_Gene_ID') %>%
    filter(Ensembl_Gene_ID %in% GSE10797_chr16) %>%
    filter(Ensembl_Gene_ID %in% common_genes) %>%
    pivot_longer(cols=-Ensembl_Gene_ID,
                 names_to='Sample_ID',
                 values_to='Expression_Level')%>%
    mutate(Dataset = 'GSE10797')

GSE59772_chr16_common = GSE59772_exp %>%
    as_tibble(rownames = 'Ensembl_Gene_ID') %>%
    filter(Ensembl_Gene_ID %in% GSE59772_chr16) %>%
    filter(Ensembl_Gene_ID %in% common_genes) %>%
    pivot_longer(cols=-Ensembl_Gene_ID,
                 names_to='Sample_ID',
                 values_to='Expression_Level')%>%
    mutate(Dataset = 'GSE59772')
```

Finally, we combine the data sets using bind_rows() and read it into a ggplot.

``` {r chr16_boxplot_3}
combined_data = bind_rows(GSE10797_chr16_common, GSE59772_chr16_common) %>%
    ggplot(aes(x = Dataset, y = Expression_Level, fill = Dataset)) +
    geom_boxplot() +
    theme_bw() +
    labs(x = 'Dataset', y='Gene Expression Level', 
         title = 'Gene Expression Levels in Chromosome 16') 
combined_data
```

# Session Info 

<details> 

```{r Session Info}
utils::sessionInfo()
```

</details>  

<br>