---
title: "AnnotatedBCGEData"
author:
  - name: "Heidi Steadman"
    affiliation:
      - &id "Brigham Young University"
  - name: "Shu-Fei Lo"
    affiliation: *id
  - name: "Stephen Piccolo"
    affiliation: *id
package: "AnnotatedBCGEData"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{AnnotatedBCGEData}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, setup, echo=FALSE, include=FALSE}
pkg <- read.dcf("../DESCRIPTION", fields = "Package")[1]
library(pkg, character.only = TRUE)
```

# Abstract

`AnnotatedBCGEData` is an [R](https://r-project.org)/[Bioconductor](https://bioconductor.org) package that provides access to more than 100 gene-expression datasets generated using microarray and RNA-Sequencing technologies. We collected these datasets from public repositories, including [Gene Expression Omnibus](http://www.ncbi.nlm.nih.gov/geo/), [ArrayExpress](https://www.ebi.ac.uk/biostudies/arrayexpress), and [The Cancer Genome Atlas](https://www.cancer.gov/ccg/research/genome-sequencing/tcga). Where feasible, we obtained and reprocessed the raw data using computational pipelines and annotations that are standardized and more updated than those used to process the data originally. Additionally, we annotated the sample metadata using ontology terms from the [National Cancer Institute Thesaurus](https://evsexplore.semantics.cancer.gov/evsexplore/welcome). This was necessary because different data submitters use different terminology to describe breast-cancer patients and samples. Mapping the metadata to ontology terms makes it possible for users to search our data using these standardized terms.

# Installation instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/).

Then install `AnnotatedBCGEData` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("AnnotatedBCGEData")
```

Or install the development version:

```{r install_dev, eval = FALSE}
BiocManager::install("srp33/AnnotatedBCGEData")
```

Finally, load the package:

```{r load_example, eval = FALSE}
library(AnnotatedBCGEData)
```

# Data retrieval

Attempting to store all of the data within this package would be infeasible due to the data's size. Thus, we store the data remotely. Once a user has identified dataset(s) to work with, our package retrieves the data from a public [Zenodo](https://zenodo.org/communities/annnotatedbcgedata/records) repository. (Alternatively, if they prefer, users can access the files directly from Zenodo in TSV format.) They can also find details about the data by exploring our Zenodo repository.

# Filtering datasets by ontology terms

```{r load_packages, include = FALSE}
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyr)
library(zen4R)
library(AnnotatedBCGEData)
library(SummarizedExperiment)
```

## Searching for ontology terms

This package contains functions designed to help users filter the included datasets based on ontology terms we have associated with them. The first function that researchers typically use is `searchOntologyTerms`. Its purpose is to help users identify ontology terms with which the data have been annotated. It accepts two parameters: a string to search for and what should be searched. The second parameter allows the user to indicate whether the string is expected to match the "Name" (default), "Definition", "URI", or "Code" of the ontology term(s). The function returns a tibble with information about the matching ontology terms. The user reviews the matching term(s) and retrieves the corresponding ontology code(s) from the tibble; these can then be passed to the `searchForDatasetsByField` or `searchForDatasetsByValue` functions.

In the following example, we are using the `searchOntologyTerms` function to search for ontology terms that match the string, "Progesterone Receptor Status". Because we did not specify a value for the `term_type` argument, the default of "Name" is used.

```{r PR_example}
searchOntologyTerms("Progesterone Receptor Status")
```

This function returns a tibble containing details about ontology terms having a name that matches the search string.

In this similar example, we are searching for ontology terms that match the shorter (and lowercase) string, "receptor status".

```{r RS_example}
searchOntologyTerms("receptor status")
```

The function returns a tibble containing details about ontology terms related to both progesterone receptor status and estrogen receptor status.

The National Cancer Institute Thesaurus provides synonyms for most terms. When a user searches by "Name" with the `searchOntologyTerms` function, it automatically checks against synonyms as well. This is helpful when the search text does not match the "preferred name" of an ontology term but instead matches a synonym. For example, we could use the following code to search for ontology terms related to progesterone receptor status using a synonym that is commonly used in medical contexts.

```{r P_example}
searchOntologyTerms("PgR Status")
```

Another alternative is to search the text within ontology term *definitions*. In this example, we are searching for definitions matching the string, "immunohistochemical". Searching by definitions is helpful when you don’t know the exact term but understand the concept you’re looking for. It lets you find the right match based on meaning rather than specific words.

```{r P_example}
searchOntologyTerms("immunohistochemical", term_type = "Definition")
```

In this case, the function returns a tibble containing details about ontology terms for concepts that use immunohistochemical methods for profiling. This includes terms related to progesterone receptor status as well as other cell-surface markers.

## Identifying datasets associated with specific ontology terms

After we have identified ontology terms for the research topic we are interested in, we can identify datasets associated with those terms. The `searchForDatasets` function accepts two parameters: ontology term code(s) and where to search for mappings between the code(s) and the data. The codes are unique identifiers associated with ontology terms (typically found using the `searchOntologyTerms` function). For the second parameter, the options are "field" (default) or "value". When "field" is specified, the function searches for fields (column names) that have been mapped to the specified code(s). When "value" is specified, the function searches for particular values in any column in a dataset that have been mapped to the specified code(s). The `searchForDatasets` function returns a tibble with metadata about data sets that are associated with the ontology code(s). The names of these datasets can then be passed to the `getBCGEData` function to retrieve the actual data.

In the previous examples, we searched for ontology terms related to progesterone receptor status. One of the matching terms had a name of "Progesterone Receptor Status" and a code of "C16149". In the following example, we are searching for datasets with a column (field) name that has been mapped to this ontology term. This will help us identify datasets where the researchers who shared the data have recorded this information. We specify the term's code when invoking the `searchForDatasets` function. We leave the `term_type` parameter out because the default is "field".

```{r PR_example_p2}
searchForDatasets("C16149")
```

```{r PR_example_p3}
searchForDatasets("C15496", term_type = "value")
```

## Filtering data by dataset-level metadata

In addition to the ontology, users can also decide which data sets to use based on the metadata about the collection of the data. This is accessible with the function `getDatasetMetadata` and passing a vector of data set IDs. Below is an example of using this function.

```{r get_datameta}
getDatasetMetadata(c("GSE41197", "GSE59772", "GSE10797"))
```

# Making SummarizedExperiment Objects

Downloading this package does not download the data sets onto the user's machine. To load the SummarizedExperiment object for a data set included in AnnotatedBCGEData, the user chooses the name of the data set and passes it as string to the function getBCGEData. This function downloads the data and creates the SummarizedExperiment object that is accessible to the user. <br>
<br>
The SummarizedExperiment object for each data set includes 3 matrices. The first, which we load as expression_data, is a matrix with the samples as columns and genes as rows. The Ensembl gene ID is used as the row names. The data in the actual matrix is the expression level of the gene for the sample. <br>
<br>
The second matrix we load as sample_metadata. This is information about the samples themselves, such as demographics for the patient the sample came from. The samples are rows and the information is in the columns. <br>
<br>
The third matrix we load as feature_data. This is the information about the genes that were tested for each sample. This includes information like the chromosome the gene is located on, the Entrez gene ID, and the gene name. The genes are the rows and the information about the samples is in the columns.

## getBCGEData
The `getBCGEData` function returns the SummarizedExperiment object for the requested data set. The function creates a cache (using BiocFileCache) on the user's machine. All downloaded files are saved to this cache so that files do not need to be redownloaded every time the R session is restarted. By default, the most recent version of the data that is found in the Zenodo repository is downloaded. When called for the first time, the user will be prompted to select or enter a location for the cache to be saved.<br> 
<br>
We understand the importance of data versioning and reproducibility. For that reason, there is additional functionality included in the getBCGEData function that will allow the user to specify which version of the data they want to download or use. The argument `v` to the function takes an integer that is used to find the specified version. If that version does not exist or needs to be downloaded, the user will see a message in the R console. If `v` is not specified, the most recent version is downloaded and used. <br>
<br>
While the actual files are stored in the AnnotatedBCGEData cache that is created, for reasons of data version checks, the SummarizedExperiment objects are not stored in the cache and need to be created again when the R session is restarted. This is a fairly quick process as the data already exists. This requires nothing more than calling `getBCGEData` on the desired data set at the beginning of every session.<br>
<br>
The following is a simple example of creating an object and accessing the data.

```{r example, eval = requireNamespace('AnnotatedBCGEData')}
GSE41197_SE = getBCGEData("GSE41197")
GSE41197_expression_data = assay(GSE41197_SE)
sample_metadata = colData(GSE41197_SE)
feature_data = rowData(GSE41197_SE)
```

TODO: if the user specifies a filepath to save the data to when calling `getBCGEData`. If no filepath is specified, the data will be saved to a temporary directory that will refresh when the R session is restarted.
TODO: The data can optionally be stored in a persistent cache on the user's machine. This caching process avoids having to re-download the data across R sessions.
TODO: At the time of package development, the Zenodo API limits the number of requests per minute to 30. Because of this, we have found it necessary to allow breaks between successive calls to `getBCGEData`. The API limitations should not cause problems for the user, but users should be aware that this may cause delays when downloading large amounts of data.

Similarly, if the user wanted to access an older version of the same dataset, the function could be used as follows;
``` {r version_example}
GSE41197_v2 = getBCGEData("GSE41197", v=2)

GSE41197_v2_expressions = assay(GSE41197_v2)
v2_metadata = colData(GSE41197_v2)
v2_feature_data = rowData(GSE41197_v2)
```
## Example 1: Plotting Expression Values

In the following example, we demonstrate data exploration that can be done easily with the package. We first access the matrix of gene expression values. To make it easier to analyze, we convert it into a tibble and select the first 10 rows. We then used ggplot() to create a bar plot showing the gene expression levels for one sample across 10 genes.

```{r plot_example_part_2}
exp_tib = as_tibble(GSE41197_expression_data, 
                    rownames='Ensembl_Gene_ID')[1:10,] %>%
    ggplot(aes(x=Ensembl_Gene_ID, y=GSM1010328)) +
    geom_col() +
    labs(x='Ensembl Gene ID',y='Sample ID GSM1010328') +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

exp_tib
```

## Example 2: Using Metadata

In this example, we combine using both the metadata and the expression data to create a heat map of gene expression levels.<br>
First, we load the data sets, turn them into tibbles, and extract both the metadata and expression data. We then filter the expression to only include the samples that are identified as stroma cells. Finally, we create two vectors with all the gene names in each tibble to set them up for comparison. 

``` {r heat_map, message=FALSE, warning=FALSE}
GSE10797_SE <- getBCGEData("GSE10797")
GSE59772_SE <- getBCGEData("GSE59772")


GSE59772_exp <- assay(GSE59772_SE)
GSE10797_exp <- assay(GSE10797_SE)

GSE59772_metadata = colData(GSE59772_SE) %>%
    as_tibble(rownames = 'Sample_ID')

GSE59772_stroma_samples = filter(GSE59772_metadata, tissue == 'Stroma') %>%
    pull(Sample_ID)

GSE59772_gene_names = GSE59772_exp %>%
    as_tibble(rownames="Ensembl_Gene_ID") %>%
    pull(Ensembl_Gene_ID)


GSE10797_metadata = colData(GSE10797_SE) %>%
    as_tibble(rownames = 'Sample_ID')

GSE10797_gene_names = GSE10797_exp %>%
    as_tibble(rownames="Ensembl_Gene_ID") %>%
    pull(Ensembl_Gene_ID)

GSE10797_stroma_samples = filter(GSE10797_metadata, 
                tissue_source == 'stromal cells from breast cancer patient')%>%
    pull(Sample_ID)
```

Next, we create a vector that has only the names of the genes the data sets have in common. We use this vector to filter the expression tibbles again to only include rows with genes in the vector. 

``` {r heat_map_2}
common_genes = c()
for (gene in GSE59772_gene_names) {
    if (gene %in% GSE10797_gene_names) {
        common_genes = c(common_genes, gene)
    }
}

GSE59772_stroma_expressions = GSE59772_exp %>%
    as_tibble(rownames= 'Ensembl_Gene_ID') %>%
    select(Ensembl_Gene_ID, all_of(GSE59772_stroma_samples)) %>%
    filter(Ensembl_Gene_ID %in% common_genes)
    
GSE10797_stroma_expressions = GSE10797_exp %>%
    as_tibble(rownames = 'Ensembl_Gene_ID') %>%
    select(Ensembl_Gene_ID, all_of(GSE10797_stroma_samples)) %>% 
    filter(Ensembl_Gene_ID %in% common_genes)
```

Finally, we combine the data into one tibble called combined_data. We make a new row called variance that calculates the variance across the 5 samples. We select the top 10 rows, those with greatest variance. We then use ggplot to create a heat map of the genes with the top variance, displayed below. 

``` {r heat_map_3}
combined_data = full_join(GSE10797_stroma_expressions, 
                          GSE59772_stroma_expressions)
top_variance_genes = combined_data %>%
    rowwise() %>%
    mutate(variance = var(c_across(starts_with('GSM')))) %>%
    ungroup() %>%
    arrange(desc(variance)) %>%
    slice_head(n=10)

top_variance_map = top_variance_genes %>%
    select(-variance) %>%
    pivot_longer(cols = starts_with('GSM'), 
                 names_to = 'Sample_ID', 
                 values_to = 'Expression') %>%
    ggplot(aes(x = Sample_ID, y = Ensembl_Gene_ID, fill = Expression)) +
    geom_tile()+
    labs(x = 'Sample', y = 'Gene', fill = 'Expression', 
         title = 'Top 10 Genes in Stroma Samples with Highest Variance') + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1), 
          plot.title = element_text(hjust = 0.5, face = "bold"))
top_variance_map
```

## Example 4: Using Feature Data

In this final example, we use aspects of the feature data along with the expression data. <br> <br>

First, we get the feature data and convert it to a tibble. We then filter this data for each data set to only include genes on chromosome 16. Similar to the last example, we also make a vector with the gene names the data sets have in common. 

``` {r chr16_boxplot, warning=FALSE}
GSE59772_featuredata = rowData(GSE59772_SE) %>%
    as_tibble(rownames = 'Ensembl_Gene_ID')

GSE59772_chr16 = GSE59772_featuredata %>%
    filter(Chromosome == '16') %>%
    pull(Ensembl_Gene_ID)

GSE59772_gene_names = GSE59772_exp %>%
    as_tibble(rownames="Ensembl_Gene_ID") %>%
    pull(Ensembl_Gene_ID)


GSE10797_featuredata = rowData(GSE10797_SE) %>%
    as_tibble(rownames = 'Ensembl_Gene_ID')

GSE10797_chr16 = GSE10797_featuredata %>%
    filter(Chromosome == '16') %>%
    pull(Ensembl_Gene_ID)

GSE10797_gene_names = GSE10797_exp %>%
    as_tibble(rownames="Ensembl_Gene_ID") %>%
    pull(Ensembl_Gene_ID)

common_genes = c()
for (gene in GSE59772_gene_names) {
    if (gene %in% GSE10797_gene_names) {
        common_genes = c(common_genes, gene)
    }
}
```

Next, we filter the expression data to include only genes on chromosome 16, then filter again to only include genes the data sets have in common. We pivot and mutate to make the data ready for the boxplot.

``` {r chr16_boxplot_2}
GSE10797_chr16_common = GSE10797_exp %>%
    as_tibble(rownames = 'Ensembl_Gene_ID') %>%
    filter(Ensembl_Gene_ID %in% GSE10797_chr16) %>%
    filter(Ensembl_Gene_ID %in% common_genes) %>%
    pivot_longer(cols=-Ensembl_Gene_ID,
                 names_to='Sample_ID',
                 values_to='Expression_Level')%>%
    mutate(Dataset = 'GSE10797')

GSE59772_chr16_common = GSE59772_exp %>%
    as_tibble(rownames = 'Ensembl_Gene_ID') %>%
    filter(Ensembl_Gene_ID %in% GSE59772_chr16) %>%
    filter(Ensembl_Gene_ID %in% common_genes) %>%
    pivot_longer(cols=-Ensembl_Gene_ID,
                 names_to='Sample_ID',
                 values_to='Expression_Level')%>%
    mutate(Dataset = 'GSE59772')
```

Finally, we combine the data sets using bind_rows() and read it into a ggplot.

``` {r chr16_boxplot_3}
combined_data = bind_rows(GSE10797_chr16_common, GSE59772_chr16_common) %>%
    ggplot(aes(x = Dataset, y = Expression_Level, fill = Dataset)) +
    geom_boxplot() +
    theme_bw() +
    labs(x = 'Dataset', y='Gene Expression Level', 
         title = 'Gene Expression Levels in Chromosome 16') 
combined_data
```

# Session Info 

<details> 

```{r Session Info}
utils::sessionInfo()
```

</details>  

<br>